---
title: "Compare Q4 on C/N 116 DR 0.06"
author: "Simone Zaghen"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: 
  html_document: 
    toc: true
    number_section: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../results/Reports") })
---

# Load required packages

```{r, results='hide', message=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(piano)
library(ggpubr)
library(EnhancedVolcano)

```

# Load data

```{r}
load(file = "../results/Output_Preprocessing/x_Filtered_2_TMM.RData")
load(file = "../results/Output_Preprocessing/gene_counts.RData")
load(file = "../results/Output_Preprocessing/lcpm_x_Filtered_2_TMM.RData")

```

# DEG on interesting conditions

```{r}
Subset <- x_Filtered_2_TMM[,which(x_Filtered_2_TMM$samples$CN_Ratio == "116" &
                                  x_Filtered_2_TMM$samples$NitrogenSource == "Ammonium Sulphate" &
                                  x_Filtered_2_TMM$samples$DR == "0.06")]

```

## Check for outliers 

```{r warning=FALSE, fig.keep='last'}
mds <- plotMDS(Subset)

toplot <- data.frame(Dim1 = mds$x, Dim2 = mds$y, Strain = factor(Subset$samples$Strain), DR = factor(Subset$samples$DR))
mds$var.explained <- round(mds$var.explained*100, 0)
ggplot(toplot, aes(Dim1, Dim2, colour = Strain)) + geom_point() +
   xlab(paste0("PC1: ",mds$var.explained[1],"% variance")) +
   ylab(paste0("PC2: ",mds$var.explained[2],"% variance")) +
   geom_text(aes(label = Strain), position = position_nudge(x = 0.2), show.legend = FALSE) + 
   theme_bw()

```

## Create design matrix

```{r}
#Define levels for design matrix
group <- Subset$samples$Strain
  
#Design model matrix
design <- model.matrix(~ 0 + group, data = Subset) #define design matrix
rownames(design) <- rownames(Subset$samples)
colnames(design) <- gsub("group", "", colnames(design))
design #check the design matrix

```

## Create contrast matrix

```{r}
contr.matrix <- makeContrasts(
   JFYL007vsOKYL029 = JFYL007 - OKYL029,
   OKYL049vsOKYL029 = OKYL049 - OKYL029,
   levels = colnames(design))
contr.matrix

```

## Voom transformation

```{r}
v <- voom(Subset, design, plot = TRUE) #voom transform

```

## Fit the model

```{r}
#Linear modelling and empirical Bayes moderation
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix) 
efit <- eBayes(vfit)
plotSA(efit)

```


```{r fig.height=4, fig.width=10}
summary(decideTests(efit))

par(mfrow = c(1,2))
for (i in 1:ncol(contr.matrix)) {
   plotMD(efit, column = i)
}

```

## Check p value distribution

```{r fig.height=4,fig.width=10}

#move the results of the comparisons in a list
comp_list <- list()
for (i in 1:ncol(contr.matrix)) {
   name = colnames(contr.matrix)[i]
   tmp = topTreat(efit, coef = i, n = Inf) 
   comp_list[[name]] <- tmp
}

#plot the p value distribution
par(mfrow = c(1,2))
for (i in 1:length(comp_list)) {
   name = names(comp_list[i])
   hist(as.vector(comp_list[[i]][[5]]), main = name, xlab = "P-Value")
}

if (!dir.exists("../results/Comparison_Output")){dir.create("../results/Comparison_Output")}
save(comp_list, file = "../results/Comparison_Output/comp_list_DR006.RData")

```

## Volcano plot

```{r, fig.width=8, fig.height=6, warning=FALSE}
for (i in 1:length(comp_list)) {
        plot <- EnhancedVolcano(comp_list[[i]],
                        lab = NA,
                        x = "logFC",
                        y = "P.Value",
                        pCutoff = 0.01,
                        FCcutoff = 1,
                        ylim = c(0,15),
                        title = paste0(names(comp_list[i])),
                        subtitle = NULL,
                        xlim = c(-6,6),
                        )
  print(plot)
}


```


# Gene set analysis

## Tidy and import data

```{r, warning=FALSE}

rm(list=ls()[!(ls() %in% c("comp_list"))]) #clean the working environment

source('../code/Functions/consGSA.R')
source('../code/Functions/consGSAplot.R')

```

## Create gene set collection
Create gene set collection and subset the analysis only to different levels of the GSC. 

```{r}
gene_sets <- read.csv2("../data/Gene_Set.csv") #import the gene set collection

#We have three levels of analysis, and we want to perform a GSA for each
#We need to split the GSA by level
gene_sets <- list(geneSetBP = loadGSC(gene_sets %>%
                                      filter(Level %in% "biological_process") %>%
                                      select(c("genes", "Term")),
                                      type = "data.frame"),
                  geneSetMF = loadGSC(gene_sets %>%
                                      filter(Level %in% "molecular_function") %>%
                                      select(c("genes", "Term")),
                                      type = "data.frame"),
                  geneSetCC = loadGSC(gene_sets %>%
                                      filter(Level %in% "cellular_component") %>%
                                      select(c("genes", "Term")),
                                      type = "data.frame"))
```

## Perform GSA
Loop the consensus gene set analysis through all the comparisons. 

```{r, message=FALSE, warning=FALSE, results='hide'}
#create directories where to save output of GSA
if (!dir.exists("../results/GSA/")){dir.create("../results/GSA/")}
if (!dir.exists("../results/GSA/DR_006")){dir.create("../results/GSA/DR_006")}

GSA_results <- list() #Create list to store results of the consGSA function

for (i in 1:length(names(comp_list))) {
   TopTable <- as.data.frame(comp_list[i], col.names = "", row.names = 1) #create df for GSA
     GSA_results[[i]] <- list(
       BP = consGSA(TopTable, gene_sets$geneSetBP), #run consensus GSA for biological process
       MF = consGSA(TopTable, gene_sets$geneSetMF), #run consensus GSA for molecular function
       CC = consGSA(TopTable, gene_sets$geneSetCC)) #run consensus GSA for cellular component
   names(GSA_results)[i] <- names(comp_list)[i] #assign names to each element of the list
}

```


## Plot results GSA

```{r, warning=FALSE, results='hide', message=FALSE, fig.height=15, fig.width=10}
path <- "../results/GSA/DR_006/" #where to save graphs
Pcutoff <- 0.01 #set p value cutoff
consGSA_plot_data <- list() #create list in which to save the graphs

#generate the graphs and save them in the list, along with the data to generate them
for (i in 1:length(GSA_results)) { #loop through all the GSAs we have
  consGSA_plot_data[[i]] <- list( #the results will be saved in a list
    BP = consGSAplot(resList = GSA_results[[i]][[1]], #create the consensous plot for BP
                     rankScore = 25, 
                     Pcutoff = Pcutoff, 
                     distinct = 'distinct', 
                     title = paste0("Biological Process - rankScore 25", " ,pval < ", Pcutoff)),
    
    MF = consGSAplot(resList =  GSA_results[[i]][[2]], #create the consensous plot for MF
                     rankScore = 25, 
                     Pcutoff = Pcutoff, 
                     distinct = 'distinct', 
                     title = paste0("Molecular Function - rankScore 25" , " ,pval < ", Pcutoff)),
    CC = consGSAplot(resList = GSA_results[[i]][[3]], #create the consensous plot for CC
                     rankScore = 25, 
                     Pcutoff = Pcutoff, 
                     distinct = 'distinct', 
                     title = paste0("Cellular Component - rankScore 25", " ,pval < ", Pcutoff)))
  
  names(consGSA_plot_data)[i] <- names(GSA_results)[i] #assign the names to the list of graphs
  
    #For each comparison arrange the three (BP, MF,CC) graphs and print the final plot
    print(ggarrange(consGSA_plot_data[[i]][["BP"]][["plot"]], 
                        consGSA_plot_data[[i]][["MF"]][["plot"]], 
                        consGSA_plot_data[[i]][["CC"]][["plot"]], 
                        ncol=1, 
                        nrow=3, 
                        common.legend = TRUE, 
                        legend = "bottom") %>%
      annotate_figure(top = text_grob((names(comp_list)[i]), 
                                       color = "red", face = "bold", size = 18))) %>%
      ggsave(path = path, 
           file=paste0("plot_", names(comp_list[i]), ".pdf"), width = 10, height = 16)
}
```

Save GSA results as R environment and load them in case of further analysis - avoid re-running GSA every time.

```{r}
save(GSA_results, file = "../results/GSA/DR_006/gsaRes_3.RData")
#load(file = "../results/GSA/DR_006/gsaRes_3.RData")

save(consGSA_plot_data, file = "../results/GSA/DR_006/consGSA_plot_data_3.RData")
#load(file = "../results/GSA/DR_006/consGSA_plot_data_3.RData")

```
