---
title: "Heatmap of GO terms"
author: "Simone Zaghen"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: 
  html_document: 
    toc: true
    number_section: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../results/Reports") })
---

# Load packages
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(gplots)

```

# Preparing matrices for heatmaps 
## Import data
We first import only the Biological Process (BP) data. Later the same code will run for the other two levels of the GO term analysis (cellular component (CC) and molecular function (MF)).

```{r, warning=FALSE}
setwd("../results/Gene_Set_Analysis/OutputGSA/")  
comparison.list.BP <- list.files(pattern = "BP_*", full.names = FALSE) #file names of excel files (comparison outputs) to be imported
comparison.list.CC <- list.files(pattern = "CC_*", full.names = FALSE) #file names of excel files (comparison outputs) to be imported
comparison.list.MF <- list.files(pattern = "MF_*", full.names = FALSE) #file names of excel files (comparison outputs) to be imported

BP.list <- sapply(comparison.list.BP, read.csv2, simplify = FALSE) #import comparison outputs
CC.list <- sapply(comparison.list.CC, read.csv2, simplify = FALSE) #import comparison outputs
MF.list <- sapply(comparison.list.MF, read.csv2, simplify = FALSE) #import comparison outputs

comparison.list.BP <- gsub("*BP_", "", comparison.list.BP)
comparison.list.BP <- gsub(".csv", "", comparison.list.BP)
names(BP.list) <- comparison.list.BP
BP.list = map(BP.list, ~ filter(.x, !(geneset %in% "All"))) #select only relevant columns

comparison.list.CC <- gsub("*CC_", "", comparison.list.CC)
comparison.list.CC <- gsub(".csv", "", comparison.list.CC)
names(CC.list) <- comparison.list.CC
CC.list = map(CC.list, ~ filter(.x, !(geneset %in% "All"))) #select only relevant columns

comparison.list.MF <- gsub("*MF_", "", comparison.list.MF)
comparison.list.MF <- gsub(".csv", "", comparison.list.MF)
names(MF.list) <- comparison.list.MF
MF.list = map(MF.list, ~ filter(.x, !(geneset %in% "All"))) #select only relevant columns

```

## Create matrices for biological process (BP)
We first create an empty matrix with the GO terms as rownames and the samples as column names. Then the matrix is filled with 1 if the GO term is present in the sample, or it is filled with a zero if it is not. 

```{r}
GO_terms_BP = '' #create an empty vector

# go through the elements of the list and extract the GO terms from "geneset"
for (i in 1:length(comparison.list.BP)) {
      GO_terms_BP <- append(GO_terms_BP, BP.list[[comparison.list.BP[i]]][["geneset"]])
}

GO_terms_BP <- unique(GO_terms_BP) #only keep unique GO terms
GO_terms_BP <- GO_terms_BP[2:length(GO_terms_BP)] #remove the first empty entry of the vector

#create empty matrix to fill and assign row and column names
Biological_Process <- matrix(nrow = length(GO_terms_BP), ncol = length(comparison.list.BP))
rownames(Biological_Process) <- GO_terms_BP
colnames(Biological_Process) <- comparison.list.BP

for (i in 1:length(GO_terms_BP)) {
  for (j in 1: length(comparison.list.BP)) {
    if (GO_terms_BP[i] %in% BP.list[[comparison.list.BP[j]]][["geneset"]]) {
      position = which(BP.list[[comparison.list.BP[j]]][[2]]==GO_terms_BP[i])
      Biological_Process[i,j] = BP.list[[comparison.list.BP[j]]][["rank"]][position]
      } 
    else{Biological_Process[i,j] = 0
    }
  }
}

```

## Create matrices for cellular component (CC)
The previous steps are repeated for CC.
```{r, warning=FALSE}

GO_terms_CC = '' #create an empty vector

# go through the elements of the list and extract the GO terms from "geneset"
for (i in 1:length(comparison.list.CC)) {
      GO_terms_CC <- append(GO_terms_CC, CC.list[[comparison.list.CC[i]]][["geneset"]])
}

GO_terms_CC <- unique(GO_terms_CC) #only keep unique GO terms
GO_terms_CC <- GO_terms_CC[2:length(GO_terms_CC)] #remove the first empty entry of the vector

#create empty matrix to fill and assign row and column names
Cellular_Component <- matrix(nrow = length(GO_terms_CC), ncol = length(comparison.list.CC))
rownames(Cellular_Component) <- GO_terms_CC
colnames(Cellular_Component) <- comparison.list.CC

#fill the matrix: goes through the lists and puts a 1 if it can find the GO term, else a 0. 
for (i in 1:length(GO_terms_CC)) {
  for (j in 1: length(comparison.list.CC)) {
    if (GO_terms_CC[i] %in% CC.list[[comparison.list.CC[j]]][["geneset"]]) {
      position = which(CC.list[[comparison.list.CC[j]]][[2]]==GO_terms_CC[i])
      Cellular_Component[i,j] = CC.list[[comparison.list.CC[j]]][["rank"]][position]
      } 
    else{Cellular_Component[i,j] = 0
    }
  }
}

```

## Create matrices for molecular function (MF)
The previous steps are repeated for MF. 
```{r, warning=FALSE}

GO_terms_MF = '' #create an empty vector

# go through the elements of the list and extract the GO terms from "geneset"
for (i in 1:length(comparison.list.MF)) {
      GO_terms_MF <- append(GO_terms_MF, MF.list[[comparison.list.MF[i]]][["geneset"]])
}

GO_terms_MF <- unique(GO_terms_MF) #only keep unique GO terms
GO_terms_MF <- GO_terms_MF[2:length(GO_terms_MF)] #remove the first empty entry of the vector

#create empty matrix to fill and assign row and column names

Molecular_Function <- matrix(nrow = length(GO_terms_MF), ncol = length(comparison.list.MF))
rownames(Molecular_Function) <- GO_terms_MF
colnames(Molecular_Function) <- comparison.list.MF

#fill the matrix: goes through the lists and puts a 1 if it can find the GO term, else a 0. 
for (i in 1:length(GO_terms_MF)) {
  for (j in 1: length(comparison.list.MF)) {
    if (GO_terms_MF[i] %in% MF.list[[comparison.list.MF[j]]][["geneset"]]) {
      position = which(MF.list[[comparison.list.MF[j]]][[2]]==GO_terms_MF[i])
      Molecular_Function[i,j] = MF.list[[comparison.list.MF[j]]][["rank"]][position]
      } 
    else{Molecular_Function[i,j] = 0
    }
  }
}

```

# Subset based on groups

## Assign samples to the subsets

```{r, warning=FALSE}
#set a rank cutoff: exclude gene sets for which no samples is below the threshold
rank_cutoff_high <- 25 
rank_cutoff_low <- 1

#List of GO levels
GO_levels <- list(Biological_Process = Biological_Process, 
                 Cellular_Component = Cellular_Component, 
                 Molecular_Function = Molecular_Function)

matrix_list <- list() #generate list in which to save output of the loop.

for (i in 1:length(GO_levels)){
    filename = paste0(names(GO_levels[i])) #generate the filename
    x = GO_levels[[i]] #select the matrix to filter
    tmp = x[rowSums(x > rank_cutoff_low & x < rank_cutoff_high) > 1,] #filter the matrix 
    matrix_list[[filename]] <- tmp #saves the filtered matrix in the matrix_list file
  }

#set order of comparisons in heatmap
col.order <- c("OKYL029_U_116_0.1vsOKYL029_AS_116_0.1",
               "OKYL049_U_116_0.1vsOKYL049_AS_116_0.1",
               "JFYL007_U_116_0.1vsJFYL007_AS_116_0.1",
               "OKYL029_U_3_0.1vsOKYL029_AS_3_0.1",
               "OKYL049_U_3_0.1vsOKYL049_AS_3_0.1",
               "JFYL007_U_3_0.1vsJFYL007_AS_3_0.1")

#order the columns in matrixes for deciding plotting order in heatmaps
for (i in 1:length(matrix_list)) {
  matrix_list[[i]] <- matrix_list[[i]][,col.order]
}


```

# Plotting
## Heatmap with RankScore as Key

The following section exports the heatmaps in a PDF files.

```{r, warning=FALSE}
if (!dir.exists("../results/Gene_Set_Analysis/Heatmaps")){dir.create("../results/Gene_Set_Analysis/Heatmaps")}
setwd("../results/Gene_Set_Analysis/Heatmaps/") 
color.palette <- c("grey90", colorRampPalette(c("red", "yellow", "white"))(n=599))

for (i in 1:length(matrix_list)) {
  filename = names(matrix_list[i]) #generate the filename of the heatmap
  pdf(paste(filename, ".pdf"), width = 10, height = 20) #open the saving device
  heatmap.2(matrix_list[[i]], #heatmap function and its setting
          Colv = FALSE,
          dendrogram = "row",
          margins = c(32, 24),
          col = color.palette, 
          main = paste(names(matrix_list[i])), 
          density.info = "none", 
          trace = "none", 
          colsep=1:nrow(matrix_list[[i]]), 
          rowsep=1:nrow(matrix_list[[i]]), 
          sepcolor = "grey20",
          sepwidth = c(0.01,0.01), 
          key=FALSE, 
          keysize = 0.2)
  dev.off() #close the saving device
}

```

# Export GO terms of each heatmap
Export all the GO terms that are used inside each heatmap.

```{r}
GO_term_heat_list <- list()

for (i in 1:length(matrix_list)) {
  v <- rownames(matrix_list[[i]]) #extract the GO term names
  GO_term_heat_list[[i]] = v #move names to list
  names(GO_term_heat_list)[i] = names(matrix_list)[i] #extract name of group comparison
}

save(GO_term_heat_list, file = "../results/Gene_Set_Analysis/Heatmaps/GO_term_heat_list.RData")

```



